---
title: "HMP 669 Project - Analysis"
author: "Group 4"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Load libraries

```{r}
library(ggplot2)
library(plyr)
library(reshape2)
library(knitr)
library(binom) 
library(purrr)
library(forcats)
library(dplyr)
```

## Load data

```{r}
omega = read.csv("omega.csv")
omega_cardio = read.csv("omega_cardio.csv")
omega_ortho = read.csv("omega_ortho.csv")
omega_other = read.csv("omega_other.csv")
omega_serv = read.csv("omega_serv.csv")
crit = read.csv("crit_cleaned.csv")
ome_dx = read.csv("ome_dx_cleaned.csv")
ome_px = read.csv("ome_px_cleaned.csv")
```

## I. Assessment of the overall patient population and current facility services

#### Top (10) disease entities treated by each group and number of patients in each of these disease entities

```{r}
# cardiac
cardiac_top_disease = omega_cardio %>% 
  group_by(DIAG_DESC) %>% 
  dplyr::summarise(count = n(),
                   prop = n()/nrow(omega_cardio) * 100) %>% 
  arrange(desc(count)) %>% 
  top_n(10)
kable(cardiac_top_disease, digits = 2, col.names = c("Cardiac condition", "Number of patients", "Percentage of patients"))

# orthopedic
ortho_top_disease = omega_ortho %>% 
  group_by(DIAG_DESC) %>% 
  dplyr::summarise(count = n(),
                   prop = n()/nrow(omega_ortho) * 100) %>% 
  arrange(desc(count)) %>% 
  top_n(10)
kable(ortho_top_disease, digits = 2, col.names = c("Orthopedic condition", "Number of patients", "Percentage of patients"))
```

#### Number of patients admitted for cardiac and orthopedic services

```{r}
num_pat = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(count = n(),
                   perc = n()/nrow(omega) * 100)

num_pat_co = num_pat %>% 
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(num_pat_co, digits = 2, col.names = c("Service", "Number of patients", "Percentage of patients"))

ggplot(data = num_pat, 
       mapping = aes(x = reorder(SERVICE, perc), y = perc, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Percentage of patients by service") + 
  xlab("Service") + 
  ylab("Percentage of patients") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

#### Demographics of patient groups

**Age**

```{r}
avg_age = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(avg_age = mean(AGE, na.rm = TRUE))

avg_age_co = avg_age %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(avg_age_co, digits = 2, col.names = c("Service", "Average age of patients"))

ggplot(data = avg_age, 
       mapping = aes(x = reorder(SERVICE, avg_age), y = avg_age, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Average age of patients by service") + 
  xlab("Service") + 
  ylab("Average age of patients") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$AGE), lty = 2, col = "black")
```

\newpage

**Race**

```{r}
# cardio                                                                                      
cardio_race = omega_cardio %>% 
  group_by(RACE) %>% 
  dplyr::summarise(count = n())
cardio_race$perc = cardio_race$count/nrow(omega_cardio) * 100
kable(cardio_race, digits = 2, col.names = c("Race", "Number of patients", "Percentage of patients"))

# ortho
ortho_race = omega_ortho %>% 
  group_by(RACE) %>% 
  dplyr::summarise(count = n())
ortho_race$perc = ortho_race$count/nrow(omega_ortho) * 100
kable(ortho_race, digits = 2, col.names = c("Race", "Number of patients", "Percentage of patients"))

race = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, RACE) %>%
  dplyr::summarise(count = n())
race$perc = ifelse(race$SERVICE == "Cardiology", race$count/nrow(omega_cardio) * 100, race$count/nrow(omega_ortho) * 100)

ggplot(data = race, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(RACE))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by race by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Race", lab = c("Asiatic", "Black", "Other", "Unknown", "White")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

**Gender**

```{r}
# cardio
cardio_gender = omega_cardio %>% 
  group_by(SEX) %>% 
  dplyr::summarise(count = n())
cardio_gender$perc = cardio_gender$count/nrow(omega_cardio) * 100
kable(cardio_gender, digits = 2, col.names = c("Gender", "Number of patients", "Percentage of patients"))

# ortho
ortho_gender = omega_ortho %>% 
  group_by(SEX) %>% 
  dplyr::summarise(count = n())
ortho_gender$perc = ortho_gender$count/nrow(omega_ortho) * 100
kable(ortho_gender, digits = 2, col.names = c("Gender", "Number of patients", "Percentage of patients"))

gender = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, SEX) %>%
  dplyr::summarise(count = n())
gender$perc = ifelse(gender$SERVICE == "Cardiology", gender$count/nrow(omega_cardio) * 100, gender$count/nrow(omega_ortho) * 100)

ggplot(data = gender, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(SEX))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by gender by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Gender", lab = c("Female", "Male")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

**Distance**

```{r}
dist_filt = omega_serv %>% filter(DISTANCE < 65535)
avg_dist = omega_serv %>% 
  filter(DISTANCE < 65535) %>%
  group_by(SERVICE) %>% 
  dplyr::summarise(avg_dist = mean(DISTANCE, na.rm = TRUE))

avg_dist_co = avg_dist %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(avg_dist_co, digits = 2, col.names = c("Service", "Average distance traveled to Omega"))

ggplot(data = avg_dist, 
       mapping = aes(x = reorder(SERVICE, avg_dist), y = avg_dist, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Average distance traveled to Omega by service") + 
  xlab("Service") + 
  ylab("Average distance traveled") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(dist_filt$DISTANCE), lty = 2, col = "black")
```

\newpage

#### Relevant admission/discharge data

**Admission**

```{r}
# cardio
cardio_adm = omega_cardio %>% 
  group_by(ADM_SVC_DESCRIPT) %>% 
  dplyr::summarise(count = n())
cardio_adm$perc = cardio_adm$count/nrow(omega_cardio) * 100
kable(cardio_adm, digits = 2, col.names = c("Admission description", "Number of patients", "Percentage of patients"))

# ortho
ortho_adm = omega_ortho %>% 
  group_by(ADM_SVC_DESCRIPT) %>% 
  dplyr::summarise(count = n())
ortho_adm$perc = ortho_adm$count/nrow(omega_ortho) * 100
kable(ortho_adm, digits = 2, col.names = c("Admission description", "Number of patients", "Percentage of patients"))

adm = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, ADM_SVC_DESCRIPT) %>%
  dplyr::summarise(count = n())
adm$perc = ifelse(adm$SERVICE == "Cardiology", adm$count/nrow(omega_cardio) * 100, adm$count/nrow(omega_ortho) * 100)

ggplot(data = adm, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(ADM_SVC_DESCRIPT))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by admission description by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Admission description", lab = c("General Surgery", "Gynecology", "Internal Medicine", "Pediatrics", "Rehabilitation")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

\newpage

**Discharge**

```{r}
# cardio
cardio_disch = omega_cardio %>% 
  group_by(DIS_SVC_DESCRIPT) %>% 
  dplyr::summarise(count = n())
cardio_disch$perc = cardio_disch$count/nrow(omega_cardio) * 100
kable(cardio_disch, digits = 2, col.names = c("Discharge description", "Number of patients", "Percentage of patients"))

# ortho
ortho_disch = omega_ortho %>% 
  group_by(DIS_SVC_DESCRIPT) %>% 
  dplyr::summarise(count = n())
ortho_disch$perc = ortho_disch$count/nrow(omega_ortho) * 100
kable(ortho_disch, digits = 2, col.names = c("Discharge description", "Number of patients", "Percentage of patients"))

disch = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, DIS_SVC_DESCRIPT) %>%
  dplyr::summarise(count = n())
disch$perc = ifelse(disch$SERVICE == "Cardiology", disch$count/nrow(omega_cardio) * 100, disch$count/nrow(omega_ortho) * 100)

ggplot(data = disch, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(DIS_SVC_DESCRIPT))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by discharge description by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Discharge description", lab = c("General Surgery", "Gynecology", "Internal Medicine", "Pediatrics", "Rehabilitation")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

\newpage

**Disposition**

```{r}
# cardio
cardio_disp = omega_cardio %>% 
  group_by(DISPOSTN_DESCRIPT) %>% 
  dplyr::summarise(count = n())
cardio_disp$perc = cardio_disp$count/nrow(omega_cardio) * 100
kable(cardio_disp, digits = 2, col.names = c("Disposition description", "Number of patients", "Percentage of patients"))

# ortho
ortho_disp = omega_ortho %>% 
  group_by(DISPOSTN_DESCRIPT) %>% 
  dplyr::summarise(count = n())
ortho_disp$perc = ortho_disp$count/nrow(omega_ortho) * 100
kable(ortho_disp, digits = 2, col.names = c("Disposition description", "Number of patients", "Percentage of patients"))

disp = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, DISPOSTN_DESCRIPT) %>%
  dplyr::summarise(count = n())
disp$perc = ifelse(disp$SERVICE == "Cardiology", disp$count/nrow(omega_cardio) * 100, disp$count/nrow(omega_ortho) * 100)

ggplot(data = disp, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(DISPOSTN_DESCRIPT))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by disposition description by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Disposition description", 
                      lab = c("Adult foster care/group home",
                              "Against medical advice",
                              "Died",
                              "Home health care or home hospice care",
                              "Home/routine",
                              "Intermediate care facility",
                              "Other facility",
                              "Skilled nursing facility",
                              "Transferred to another acute short term hospital",
                              "Transferred to LT psych from other system facility",
                              "Transferred to Omega from other system facility",
                              "Transferred to physician rehab other than Omega",
                              "Transferred to rehab from other system facility")) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

#### LOS information

```{r}
avg_los = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(avg_los = mean(LOS, na.rm = TRUE))

avg_los_co = avg_los %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(avg_los_co, digits = 2, col.names = c("Service", "Average LOS"))

ggplot(data = avg_los, 
       mapping = aes(x = reorder(SERVICE, avg_los), y = avg_los, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Average LOS by service") + 
  xlab("Service") + 
  ylab("Average LOS") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$LOS), lty = 2)
```

## II. Intensity/utilization indicators

#### Reputation: how far do these patients travel to receive cardiac and orthopedic services compared to other services?

```{r}
dist_filt = omega_serv %>% filter(DISTANCE < 65535)
avg_dist = omega_serv %>% 
  group_by(SERVICE) %>% 
  filter(DISTANCE < 65535) %>%
  dplyr::summarise(avg_dist = mean(DISTANCE, na.rm = TRUE))

avg_dist_co = avg_dist %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(avg_dist_co, digits = 2, col.names = c("Service", "Average distance traveled to Omega"))

ggplot(data = avg_dist, 
       mapping = aes(x = reorder(SERVICE, avg_dist), y = avg_dist, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Average distance traveled to Omega by service") + 
  xlab("Service") + 
  ylab("Average distance traveled") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(dist_filt$DISTANCE), lty = 2)
```

#### Death rate

**Within 48 hours**

```{r}
omega_serv$DIED_WI_48 = ifelse(omega_serv$DIED_WI_48 == 1, 1, 0)

dw48 = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(dw48 = mean(DIED_WI_48, na.rm = TRUE)*100)

dw48_co = dw48 %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(dw48_co, digits = 2, col.names = c("Service", "Death rate within 48 hours"))

ggplot(data = dw48, 
       mapping = aes(x = reorder(SERVICE, dw48), y = dw48, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Death rate within 48 hours by service") + 
  xlab("Service") + 
  ylab("Death rate within 48 hours") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$DIED_WI_48, na.rm = TRUE)*100, lty = 2)
```

**In OR**

```{r}
omega_serv$DIED_IN_OR = ifelse(omega_serv$DIED_IN_OR == 1, 1, 0)

dor = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(dor = mean(DIED_IN_OR, na.rm = TRUE)*100)

dor_co = dor %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(dor_co, digits = 2, col.names = c("Service", "Death rate in OR"))

ggplot(data = dor, 
       mapping = aes(x = reorder(SERVICE, dor), y = dor, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Death rate in OR by service") + 
  xlab("Service") + 
  ylab("Death rate in OR") + 
  scale_fill_manual(values = c(rep("black", 3), "red", rep("black", 13), "dodgerblue", rep("black", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

**Post-operative**

```{r}
omega_serv$DIED_PSTOP = ifelse(omega_serv$DIED_PSTOP == 1, 1, 0)

pstop = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(pstop = mean(DIED_PSTOP, na.rm = TRUE)*100)

pstop_co = pstop %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(pstop_co, digits = 2, col.names = c("Service", "Post-operative death rate"))

ggplot(data = pstop, 
       mapping = aes(x = reorder(SERVICE, pstop), y = pstop, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Post-operative death rate by service") + 
  xlab("Service") + 
  ylab("Post-operative death rate") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$DIED_PSTOP, na.rm = TRUE)*100, lty = 2)
```

**On arrival**

```{r}
omega_serv$DOA = ifelse(omega_serv$DOA == 1, 1, 0)

doa = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(doa = mean(DOA, na.rm = TRUE)*100)

doa_co = doa %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(doa_co, digits = 2, col.names = c("Service", "Death rate on arrival"))

ggplot(data = doa, 
       mapping = aes(x = reorder(SERVICE, doa), y = doa, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Death upon arrival rate by service") + 
  xlab("Service") + 
  ylab("Death upon arrival rate") + 
  scale_fill_manual(values = c(rep("black", 3), "red", rep("black", 13), "dodgerblue", rep("black", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

**in ER**

```{r}
omega_serv$ER_DEATH = ifelse(omega_serv$ER_DEATH == 1, 1, 0)

er = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(er = mean(ER_DEATH, na.rm = TRUE)*100)

er_co = er %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(er_co, digits = 2, col.names = c("Service", "Death in ER rate"))

ggplot(data = er, 
       mapping = aes(x = reorder(SERVICE, er), y = er, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Death in ER rate by service") + 
  xlab("Service") + 
  ylab("Death in ER rate") + 
  scale_fill_manual(values = c(rep("black", 3), "red", rep("black", 13), "dodgerblue", rep("black", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

**Overall**

```{r}
omega_serv$DEATH = ifelse(omega_serv$DIED_IN_OR == 1 | omega_serv$DIED_PSTOP == 1 | omega_serv$DIED_WI_48 == 1 | omega_serv$DOA == 1 | omega_serv$ER_DEATH == 1, 1, 0)

death = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(death = mean(DEATH, na.rm = TRUE)*100)

death_co = death %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(death_co, digits = 2, col.names = c("Service", "Overall death rate"))

ggplot(data = death, 
       mapping = aes(x = reorder(SERVICE, death), y = death, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Overall death rate by service") + 
  xlab("Service") + 
  ylab("Overall death rate")  + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$DEATH, na.rm = TRUE)*100, lty = 2)
```

#### Percent of encounters where service provided was teaching-related

```{r}
omega_serv$TEACH_SVC = ifelse(omega_serv$TEACH_SVC == "TEACHING", 1, 0)

teach = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(teach = mean(TEACH_SVC)*100)

teach_co = teach %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(teach_co, digits = 2, col.names = c("Service", "Percent of teaching-related encounters"))

ggplot(data = teach, 
       mapping = aes(x = reorder(SERVICE, teach), y = teach, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Percent of teaching-related encounters by service") + 
  xlab("Service") + 
  ylab("Percent of teaching-related encounters") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$TEACH_SVC, na.rm = TRUE)*100, lty = 2)
```

#### Physician/surgeon information (e.g., admitting physician, specialty type, operating surgeon)

```{r}
physician = omega_serv %>% 
  group_by(PHYS_A_SPC_DESCRIPT) %>%
  dplyr::summarise(avg.charge = mean(TOT_CHARGE, na.rm = TRUE))
phys_co = physician %>%
  filter(PHYS_A_SPC_DESCRIPT %in% c("CARDIOLOGY", "ORTHOPEDICS"))
kable(phys_co, digits = 2, col.names = c("Physician specialty", "Average total charge"))

ggplot(data = physician, 
       mapping = aes(x = reorder(PHYS_A_SPC_DESCRIPT, avg.charge), y = avg.charge, fill = PHYS_A_SPC_DESCRIPT)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Average total charge by physician specialty") + 
  xlab("Physician specialty") + 
  ylab("Average total charge") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$TOT_CHARGE, na.rm = TRUE), lty = 2)
```

#### Percent of visits requiring an external review

```{r}
omega_serv$REQ_EXT_REV = ifelse(omega_serv$EXT_REV > 0, 1, 0)

ext_rev = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(ext_rev = mean(EXT_REV)*100)

ext_rev_co = ext_rev %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(ext_rev_co, digits = 2, col.names = c("Service", "Percent of visits requiring external review"))

ggplot(data = ext_rev, 
       mapping = aes(x = reorder(SERVICE, ext_rev), y = ext_rev, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Percent of visits requiring an external review by service") + 
  xlab("Service") + 
  ylab("Percent of visits requiring an external review") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$REQ_EXT_REV, na.rm = TRUE)*100, lty = 2)
```

#### ICU/CCU information

```{r}
omega_crit = omega[,c("PAT_NO", "ADM_DATE", "DIS_DATE", "SERVICE")]
crit = merge(crit, omega_crit, by = c("PAT_NO", "ADM_DATE", "DIS_DATE"), all.x = TRUE)
crit_num = crit %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(num = n())
omega_num = omega %>% group_by(SERVICE) %>% dplyr::summarise(count = n())
crit_count = merge(crit_num, omega_num, by = "SERVICE")

crit_count$prop = crit_count$num/crit_count$count * 100
crit_count = na.omit(crit_count)

crit_count_co = crit_count %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  select("SERVICE", "prop")
kable(crit_count_co, digits = 2, col.names = c("Service", "Percent of visits with ICU stay"))
```

```{r}
ggplot(data = crit_count, 
       mapping = aes(x = reorder(SERVICE, prop), y = prop, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Percent of visits requiring ICU stay") + 
  xlab("Service") + 
  ylab("Percent of visits requiring ICU stay") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 12), "dodgerblue", rep("darkgrey", 10))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = (sum(crit_count$num)/sum(crit_count$count))*100, lty = 2)
```

```{r}
new_omega = crit %>% inner_join(omega_serv)
CCU_COUNTS = table(new_omega$ICU)/nrow(new_omega)*100

plot = new_omega %>% 
  group_by(SERVICE, ICU) %>% 
  dplyr::summarise(avg_ICU_LOS = mean(ICU_LOS)) %>% 
  arrange(desc(avg_ICU_LOS))
total_length = plot %>% group_by(SERVICE) %>% dplyr::summarise(tot = sum(avg_ICU_LOS))
plot = merge(plot, total_length, by = "SERVICE")

plot %>%
  ggplot(aes(x = reorder(SERVICE, tot), y = avg_ICU_LOS, fill = ICU)) +
  coord_flip() +
  geom_bar(stat = "identity")+
   labs(x = "Service", y = "Average ICU length of stay", title = "Average length of stay by service in the ICU")+
    theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.background = element_rect(fill = "white"),
        legend.position = "right") +
  scale_fill_discrete(name = "Unit type", lab = c("CCU", "MICU", "SICU", "Special nursery"))
```

#### Principal operations performed on each group

**Top 10 operations for each service**

```{r}
cardiac_top_op = subset(omega_cardio, !is.na(OP_DESC))
ortho_top_op = subset(omega_ortho, !is.na(OP_DESC))

# cardiac
cardiac_top_op = cardiac_top_op %>% 
  group_by(OP_DESC) %>% 
  dplyr::summarise(count = n(),
                   prop = n()/nrow(omega_cardio) * 100) %>% 
  arrange(desc(count)) %>% 
  top_n(10)
kable(cardiac_top_op, digits = 2, col.names = c("Cardiac operation", "Number of patients", "Percentage of patients"))

# orthopedic
ortho_top_op = ortho_top_op %>% 
  group_by(OP_DESC) %>% 
  dplyr::summarise(count = n(),
                   prop = n()/nrow(omega_ortho) * 100) %>% 
  arrange(desc(count)) %>% 
  top_n(10)
kable(ortho_top_op, digits = 2, col.names = c("Orthopedic operation", "Number of patients", "Percentage of patients"))
```

**Number of patients with operations**

```{r}
num_op = omega %>% 
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE) %>%
  dplyr::summarise(num_op = sum(!is.na(OP_DESC)))
num_op$prop = ifelse(num_op$SERVICE == "Cardiology", num_op$num_op/nrow(omega_cardio)*100, num_op$num_op/nrow(omega_ortho)*100)
kable(num_op, digits = 2, col.names = c("Service", "Number of patients having operations", "Percentage of patients having operations"))
```

**Average number of operations performed**

```{r}
ome_px_service = merge(ome_px, omega, by = c("PAT_NO", "ADM_DATE", "DIS_DATE"), all.x = TRUE) 
ome_px_first = ome_px_service[ome_px_service$OCCURRENCE == 1,]
avg_op = ome_px_first %>% 
  filter(!is.na(SERVICE)) %>%
  group_by(SERVICE) %>%
  dplyr::summarise(total = sum(TOT_NO_OPS),
                   count = n()) %>%
  mutate(avg_op = total/count) %>%
  select(SERVICE, avg_op)

ggplot(data = avg_op, 
       mapping = aes(x = reorder(SERVICE, avg_op), y = avg_op, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Average number of operations by service") + 
  xlab("Service") + 
  ylab("Average number of operations") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

## III. Finanical viability

#### Profitability of patients from cardiac and orthopedic services

```{r}
omega_serv$PROFIT = omega_serv$TOT_PAY - omega_serv$TOT_COST

profit = omega_serv %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(avg_prof = mean(PROFIT, na.rm = TRUE))

profit_co = profit %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(profit_co, digits = 2, col.names = c("Service", "Profitability"))

ggplot(data = profit, 
       mapping = aes(x = reorder(SERVICE, avg_prof), y = avg_prof, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Profitability by service") + 
  xlab("Service") + 
  ylab("Profitability") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(omega_serv$PROFIT, na.rm = TRUE), lty = 2)
```

#### LOS outlier rates

```{r}
omega_co = subset(omega, SERVICE %in% c("Cardiology", "Orthopedic"))
ggplot(omega_co, aes(x = SERVICE, y = LOS)) +
  geom_boxplot() + 
  theme_bw() +
  ggtitle("LOS by service: Cardiology and Orthopedic") +
  xlab("Service") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

```{r}
los_outlier_i = omega_serv %>% 
  group_by(SERVICE) %>%
  mutate(OUTLIER = ifelse(LOS > quantile(LOS, 0.75) + 1.5*IQR(LOS) | LOS < quantile(LOS, 0.25) - 1.5*IQR(LOS), 1, 0))

los_outlier = los_outlier_i %>% 
  group_by(SERVICE) %>% 
  dplyr::summarise(lor = mean(OUTLIER)*100)

los_outlier_co = los_outlier %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic"))
kable(los_outlier_co, digits = 2, col.names = c("Service", "LOS outlier rate"))

ggplot(data = los_outlier, 
       mapping = aes(x = reorder(SERVICE, lor), y = lor, fill = SERVICE)) + 
  geom_bar(stat = "identity") + 
  ggtitle("LOS outlier rate by service") + 
  xlab("Service") + 
  ylab("LOS outlier rate") + 
  scale_fill_manual(values = c(rep("darkgrey", 3), "red", rep("darkgrey", 13), "dodgerblue", rep("darkgrey", 9))) + 
  guides(fill = FALSE) + 
  coord_flip() + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept = mean(los_outlier_i$OUTLIER, na.rm = TRUE)*100, lty = 2)
```

\newpage

#### payer mix

**MAJOR PAYOR**

```{r}
cardio_major = omega_cardio %>% 
  group_by(MAJOR_PAY) %>% 
  dplyr::summarise(count = n())
cardio_major$perc = cardio_major$count/nrow(omega_cardio) * 100
kable(cardio_major, digits = 2, col.names = c("Major payor", "Number of patients", "Percentage of patients"))

ortho_major = omega_ortho %>% 
  group_by(MAJOR_PAY) %>% 
  dplyr::summarise(count = n())
ortho_major$perc = ortho_major$count/nrow(omega_ortho) * 100
kable(ortho_major, digits = 2, col.names = c("Major payor", "Number of patients", "Percentage of patients"))

major = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, MAJOR_PAY) %>%
  dplyr::summarise(count = n())
major$perc = ifelse(major$SERVICE == "Cardiology", major$count/nrow(omega_cardio) * 100, major$count/nrow(omega_ortho) * 100)

ggplot(data = major, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(MAJOR_PAY))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by major payor by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Major payor", lab = c("Blue Cross", "Medicaid", "Medicare", "Private/Commercial")) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

\newpage

**PRIMARY PAYOR**

```{r}
cardio_primary = omega_cardio %>% 
  group_by(PRIM_PAY) %>% 
  dplyr::summarise(count = n())
cardio_primary$perc = cardio_primary$count/nrow(omega_cardio) * 100
kable(cardio_primary, digits = 2, col.names = c("Primary payor", "Number of patients", "Percentage of patients"))

ortho_primary = omega_ortho %>% 
  group_by(PRIM_PAY) %>% 
  dplyr::summarise(count = n())
ortho_primary$perc = ortho_primary$count/nrow(omega_ortho) * 100
kable(ortho_primary, digits = 2, col.names = c("Primary payor", "Number of patients", "Percentage of patients"))

primary = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, PRIM_PAY) %>%
  dplyr::summarise(count = n())
primary$perc = ifelse(primary$SERVICE == "Cardiology", primary$count/nrow(omega_cardio) * 100, primary$count/nrow(omega_ortho) * 100)

ggplot(data = primary, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(PRIM_PAY))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by primary payor by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Primary payor", lab = c("Blue Cross", "Commercial", "Medicaid", "Medicare", "Private", "Workers' compensation")) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

\newpage

**SECONDARY PAYOR**

```{r}
cardio_secondary = omega_cardio %>% 
  group_by(SEC_PAY) %>% 
  dplyr::summarise(count = n())
cardio_secondary$perc = cardio_secondary$count/nrow(omega_cardio) * 100
kable(cardio_secondary, digits = 2, col.names = c("Secondary payor", "Number of patients", "Percentage of patients"))

ortho_secondary = omega_ortho %>% 
  group_by(SEC_PAY) %>% 
  dplyr::summarise(count = n())
ortho_secondary$perc = ortho_secondary$count/nrow(omega_ortho) * 100
kable(ortho_secondary, digits = 2, col.names = c("Secondary payor", "Number of patients", "Percentage of patients"))

secondary = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  group_by(SERVICE, SEC_PAY) %>%
  dplyr::summarise(count = n())
secondary$perc = ifelse(secondary$SERVICE == "Cardiology", secondary$count/nrow(omega_cardio) * 100, secondary$count/nrow(omega_ortho) * 100)

ggplot(data = secondary, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(SEC_PAY))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by secondary payor by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Secondary payor", lab = c("Blue Cross", "Commercial", "Medicaid", "Medicare", "Not recorded", "Private")) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

\newpage

#### Rates of emergent vs. elective care

```{r}
cardio_admtype = omega_cardio %>% 
  group_by(ADM_TYPE) %>% 
  filter(ADM_TYPE != 0) %>%
  dplyr::summarise(count = n()) 
cardio_admtype$perc = cardio_admtype$count/nrow(omega_cardio) * 100
kable(cardio_admtype, digits = 2, col.names = c("Admission type", "Number of patients", "Percentage of patients"))

ortho_admtype = omega_ortho %>% 
  group_by(ADM_TYPE) %>% 
  filter(ADM_TYPE != 0) %>%
  dplyr::summarise(count = n())
ortho_admtype$perc = ortho_admtype$count/nrow(omega_ortho) * 100
kable(ortho_admtype, digits = 2, col.names = c("Admission type", "Number of patients", "Percentage of patients"))

admtype = omega_serv %>%
  filter(SERVICE %in% c("Cardiology", "Orthopedic")) %>%
  filter(ADM_TYPE != 0) %>%
  group_by(SERVICE, ADM_TYPE) %>%
  dplyr::summarise(count = n())
admtype$perc = ifelse(admtype$SERVICE == "Cardiology", admtype$count/nrow(omega_cardio) * 100, admtype$count/nrow(omega_ortho) * 100)

ggplot(data = admtype, 
       mapping = aes(x = SERVICE, y = perc, fill = as.factor(ADM_TYPE))) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Percentage of patients by admission type by service") +
  xlab("Service") +
  ylab("Percentage of patients") +
  scale_fill_discrete(name = "Admission type", lab = c("Elective", "Emergent")) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```
